---
title: "Best practices for programming with ggplot2"
subtitle: "⚔<br/>with xaringan"
author: "Dewey Dunnington"
institute: "Dalhousie University"
date: "2020-01-30"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, styles.css]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
library(ggplot2)

input <- list(colour_var = "class", facet_var = "manufacturer")

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, dpi = 300, fig.height = 4)
```

```{r}
library(ggplot2)
ggplot(mpg) +
  geom_point(aes(displ, hwy, colour = class)) +
  facet_wrap(vars(manufacturer))
```

---
class: center, middle, inverse

# Using ggplot2 in Shiny

---

<!-- doesn't work in the demo but works in other browsers -->
<iframe class="shinyapp" src="https://paleolimbot.shinyapps.io/ggpkg_demo/"></iframe>

---

# Programming with ggplot2

Shiny `selectInput()` values are character vectors:

```{r}
colour_var <- "class"
facet_var <- "manufacturer"
```

Mappings and facet specifications use non-standard evaluation:

```{r}
ggplot(mpg) +
  geom_point(aes(displ, hwy, colour = class)) +
  facet_wrap(vars(manufacturer))
```

---

# Programming with ggplot2

Shiny `selectInput()` values are character vectors:

```{r}
colour_var <- "class"
facet_var <- "manufacturer"
```

Mappings and facet specifications use non-standard evaluation:

```{r}
ggplot(mpg) +
  geom_point(aes(displ, hwy, colour = !!sym(colour_var))) +
  facet_wrap(vars(!!sym(facet_var)))
```

---

# Programming with ggplot2

```{r}
plot_mpg <- function(colour_var, facet_var) {
  ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = !!sym(colour_var))) +
    facet_wrap(vars(!!sym(facet_var)))
}

plot_mpg("class", "manufacturer")
```

---

# Programming with ggplot2

```{r, error=TRUE}
plot_mpg <- function(colour_var, facet_var) {
  ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = !!sym(colour_var))) +
    facet_wrap(vars(!!sym(facet_var)))
}

plot_mpg("<none>", "manufacturer")
```

---

# Programming with ggplot2

```{r, error=TRUE}
plot_mpg <- function(colour_var, facet_var) {
  colour_mapping <- sym(colour_var)
  
  ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = !!colour_mapping)) +
    facet_wrap(vars(!!sym(facet_var)))
}

plot_mpg("<none>", "manufacturer")
```

---

# Programming with ggplot2

```{r}
plot_mpg <- function(colour_var, facet_var) {
  if (colour_var != "<none>") {
    colour_mapping <- sym(colour_var)
  } else {
    colour_mapping <- NULL
  }
  
  ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = !!colour_mapping)) +
    facet_wrap(vars(!!sym(facet_var)))
}

plot_mpg("<none>", "manufacturer")
```

---

# Programming with ggplot2

```{r, error = TRUE}
plot_mpg <- function(colour_var, facet_var) {
  if (colour_var != "<none>") {
    colour_mapping <- sym(colour_var)
  } else {
    colour_mapping <- NULL
  }
  
  ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = !!colour_mapping)) +
    facet_wrap(vars(!!sym(facet_var)))
}

plot_mpg("<none>", "<none>")
```

---

# Programming with ggplot2

```{r, error = TRUE}
plot_mpg <- function(colour_var, facet_var) {
  if (colour_var != "<none>") {
    colour_mapping <- sym(colour_var)
  } else {
    colour_mapping <- NULL
  }
  
  facet <- facet_wrap(vars(!!sym(facet_var)))
  
  ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = !!colour_mapping)) +
    facet
}

plot_mpg("<none>", "<none>")
```

---

# Programming with ggplot2

```{r, error = TRUE}
plot_mpg <- function(colour_var, facet_var) {
  if (colour_var != "<none>") {
    colour_mapping <- sym(colour_var)
  } else {
    colour_mapping <- NULL
  }
  
  if (facet_var != "<none>") {
    facet <- facet_wrap(vars(!!sym(facet_var)))
  } else {
    facet <- NULL
  }
  
  ggplot(mpg) +
    geom_point(aes(displ, hwy, colour = !!colour_mapping)) +
    facet
}
```

---

```{r}
plot_mpg("<none>", "<none>")
```

---

# Programming with ggplot2

`aes(aesthetic = NULL)` is identical to `aes()`

- Use to conditionally map a variable with minimal repeated code.
- Use `!!sym("column name as string")` to convert a string to an expression that can be used within `aes()`.

`ggplot() + NULL` is identical to `ggplot()`

- Use to conditionally add a layer/facet/scale to a plot

`ggplot() + item1 + item2` is identical to `ggplot() + list(item1, item2)`

- Use to conditinally add more than one item to a plot

If you need more than one `ggplot()` call in a function, it's a good indication that you need more than one function.

---
class: center, middle, inverse

# Using ggplot2 in packages

---

# Introducing...the {plotmpg} package!

```r
# install.packages("plotmpg")
library(plotmpg)
plot_mpg("class", "manufacturer")
```

```{r, echo=FALSE}
plot_mpg("class", "manufacturer")
```

---

# Using ggplot2 in packages

Functions in a package will be used by more than just you:

- Export a function that returns data in a form suitable for use in ggplot2 in addition to (or in place of!) a plot function

Explicit namespacing (no `library()` calls):

- Use roxygen `#' @importFrom ggplot2 <function name>` or explicit namespacing (e.g., `ggplot2::<function_name>()`)
- Datasets require explicit namespacing (e.g., `ggplot2::mpg`)
- Infix functions require importing (e.g., `#' @importFrom ggplot2 %+replace%` or assignment (e.g., `` `%+replace%` <- ggplot2::`%+replace%` ``)

R CMD check: no undefined variables:

- Use the `.data` pronoun and `#' @importFrom rlang .data` (e.g., `aes(.data$displ, .data$hwy)`))
